### 2009-05-29 response-to-reviews2.txt

> p.13 The DP inference algorithm section says "the standard CYK ... 
> algorithms fail". But then it says "Maximum likelihood inference is 
> performed with a CYK algorithm", and even "maximum likelihood 
> inference can therefore be performed with a standard three sequence 
> CYK algorithm". Which is it?

Thank you for pointing out this inconsistency, which we have clarified
in the revised text.  Using the standard CYK algorithm on the composed
model fails; however, the model can be simply modified as described in
the revised text to permit standard CYK to work properly.

> p.14 It seems possible that the problem with "null cycles" (residues 
> in ancestors that leave no homologous residues in observed sequences) 
> may simply go away if you view the problem as a missing data problem, 
> and that the likelihood being inferred is the likelihood of alignment 
> columns needed to account for the observed data (marginalized over 
> missing columns needed to account for null cycles) rather than the 
> joint likelihood of the "observed" columns and the "null cycle" 
> columns. This is what dropped out of the equations in the phylogenetic 
> inference described by Rivas and Eddy (2008) PLoS CB 4:e1000172, for 
> example.

We completely agree. Such an algorithm was initially a goal of this
work, but was dropped. We thank the reviewer for challenging us to
try again; a verbal presentation of the algorithm we came up with is
given in the revised text, with a full formal description (including
notation) in the supplement. (Since this is a standalone result that
can be applied to any SCFG, we think it fits better as a technical
appendix in the supplement, with a verbal description in the main
paper; the obvious cost is omitting some detail from the main paper.
The PLoS Authors' Guidelines (which recommend using the supplement for
algorithmic details) and the reviewer's preference for including more
detail in the manuscript are potentially in conflict on the best
approach to take here. We welcome advice on how to improve the flow
generally.)

As a technical aside, the primary difficulty with SCFG null cycle
elimination is that while this marginalization is reasonably
straightforward for independent-sites models and HMMs, essentially
involving summing a geometric series to obtain a correction factor,
the possible presence of null bifurcations complicates matters for
SCFGs.  Indeed, no *closed form* elimination is possible (due to
bifurcations, such an elimination would involve solving simultaneous
nonlinear equations, as explained in the supplement) but an iterative
approximation (e.g. Newton-Raphson) appears adequate.

@@

The 5' and 3' ends of structures are now labeled (e.g., Figures 3 and
4 of the revised manuscript).

> Figure 3: It is very unclear what this figure is supposed to be 
> representing (and I had the same problem with it when it appeared in 
> reference [28]). It appears to simply be a representation of a parse 
> tree of an individual RNA structure. It does not seem to show any kind 
> of evolutionary model from one sequence/structure/parse tree to 
> another, as the caption and text would seem to imply.

We have updated this figure and included an additional figure (also
extended from a figure in [28]) which illustrates the requested
mutations.  Both figures now include the structures for clarity.

Additionally, we have used our simulation script to prepare animations
showing illustrative evolutionary trajectories of RNA structures under
our model.  The software to produce these animations is available and
briefly mentioned in the supplement.  (If there were some way to
include these animations in the online version, we would like to do
that, though this may have to wait for the <video> tag of HTML 5...)

### 2009-05-29 paper.tex
Revised supplement; Ian's changes to Rob's changes

Our problem statement is this:
%\vspace{.5\baselineskip}
%\fbox{
%\parbox{.8\textwidth}
\textbf{Given a phylogenetic tree relating several structured RNAs and a
  description of the evolution of a structured RNA
  along a single branch of the tree (in the form of a Pair SCFG),
  (1) find the corresponding
  phylogenetic multiple-sequence grammar and (2) use that grammar to
  reconstruct, {\em a posteriori}, the evolutionary histories of the
  RNAs.}
%}
%\vspace{.5\baselineskip}
We assume that the phylogeny, including both the tree topology and
branch lengths, is known.

This paper focuses on model construction and inference algorithms
rather than the heuristics and constraints which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of general inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
Such constraints are
commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
Alignment and structural constraints can be combined \cite{Holmes2005}.

>>>>

Our problem statement is this:
%\vspace{.5\baselineskip}
%\fbox{
%\parbox{.8\textwidth}
\textbf{Given a phylogenetic tree relating several structured RNAs and a
  description of the evolution of a structured RNA
  along a single branch of the tree (in the form of a Pair SCFG),
  (1) find the corresponding
  phylogenetic multiple-sequence grammar and (2) use that grammar to
  reconstruct, {\em a posteriori}, the evolutionary histories of the
  RNAs.}
%}
%\vspace{.5\baselineskip}

We assume here that the phylogeny, including both the tree topology and branch lengths, is given.
This constraint somewhat arbitrarily confines the problem domain to the rather restricted case where we already have a phylogeny (and potentially a consensus structure),
and we seek only an alignment plus ancestral reconstruction.
It is important to note, however, that as far as the {\em model} is concerned, none of the constraints (phylogeny, structure or alignment) are final.
For example, the probabilistic framework readily allows us to compare likelihoods of two different phylogenetic trees, by constructing a composite transducer for each tree.
Thus, MCMC samplers for alignments could in principal be extended to phylogenies (albeit at a computational cost).
We revisit such MCMC samplers in the Discussion.

This paper focuses on model construction and inference algorithms
rather than the heuristics which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of general inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
Such constraints are
commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
Alignment and structural constraints can be combined \cite{Holmes2005}.
